---
title: "An Iterative Math Template for Investigating the Constraints of g,h,w,v"
author: "Xuanyu"
date: "06/18/2024"
output: html_document
---
# A gentle readme
To use this markdown, specify the starting values into the list `startingParamList1`. 
Please specify the number of condition you want to do the iterative math for in the variable `condition`. 
Also, specify the number of generations you want to run the iterative math for in the variable `gens`. 

he second chunk of code initializes the matrices and variables for the iterative math.

The third chunk of code runs the iterative math for the specified number of generations.

The constraints of g,h,w,v will be checked at the end of the markdown.

# first chunk: specify the starting values
```{r setup}
# a list of starting parameters for the iterative math
# feel free to add or change values in the list
startingParamList1 <- list(vg1 = c(.8,.6,.3,.5,.4,.35),
						   vg2 = c(.2,.7,.4,.6,.75,.2),
						   am11 = c(0.4,0.4,0.4,0.4,0.4,0.4),
						   am12 = c(0.2,0.2,0.2,0.2,0.2,0.2),
						   am21 = c(0.1,0.1,0.1,0.1,0.1,0.1),
						   am22 = c(0.3,0.3,0.3,0.3,0.3,0.3),
						   f11 = c(0.15,0.15,0.15,0.15,0.15,0.15),
						   f12 = c(0.1,0.1,0.1,0.1,0.1,0.1),
						   f21 = c(0.05,0.05,0.05,0.05,0.05,0.05),
						   f22 = c(0.1,0.1,0.1,0.1,0.1,0.1),
						   Nfam = c(rep(5e4,6)),
						   rg = rep(.1,6),
						   re = rep(.1,6),
						   prop.h2.latent1 = c(0.2,.3,.2,.1,.4,.65),
						   prop.h2.latent2 = c(0.7,.4,.3,.5,.6,.8))

# please specify the number of condition in the list you want to do the iterative math for
condition = 1
# specify the number of generations you want to run iterative math for
gens = 15
```

# second chunk: initialize the matrices and variables (don't need to tweak unless interested)
```{r initializeMatrices}

     # WILDCARD parameters
    
	pop.size <-  startingParamList1["Nfam"][[1]][[condition]] #maybe something like 2e4, or 20000, when running for real
	num.cvs <- 200 #maybe 25
	seed <- 62*condition
	num.gen <-  20 #8 should be sufficient to get to AM equilibrium
	avoid.inb <- TRUE #avoid inbreeding?
	save.covariances <- TRUE #save the covariance matrices?
	save.history <- TRUE #save the data for each generation?
	min.maf <- .1 #.2 for real
	max.maf <- .5 #.5

	#USER INPUT VARIABLES
	#VG
	vg1 <- startingParamList1["vg1"][[1]][[condition]] #trait 1 vg
	vg2 <- startingParamList1["vg2"][[1]][[condition]] #trait 2 vg
	rg <- startingParamList1["rg"][[1]][[condition]] #genetic CORRELATION @t0 bw trait 1 and trait 2 for both obs. PGS and latent PGS (assumed to be the same). NOTE: this is NOT the full rg at t0. It is the rg bw PGSs, and the rg bw LGSs. The full rg may be a bit different (Simpson's paradox)
	(k2.matrix <- matrix(c(1,rg,rg,1),nrow=2,byrow=T)) #k2 matrix is 2 * k matrix - i.e., genotypic (instead of haplotypic) var/covar at t0
	prop.h2.latent1 <- startingParamList1["prop.h2.latent1"][[1]][[condition]] #  trait 1, e.g., height
	prop.h2.latent2 <- startingParamList1["prop.h2.latent2"][[1]][[condition]] # trait 2, e.g., IQ

	#AM - these are NOT the mu copaths. They are the CORRELATIONS between male & female traits
	am11 <-  startingParamList1["am11"][[1]][[condition]] #height.m-height.f am across 2 its
	am12 <-  startingParamList1["am12"][[1]][[condition]] #height.m-iq.f; e.g., tall males choose smart females
	am21 <-  startingParamList1["am21"][[1]][[condition]] #iq.m-height.f
	am22 <-  startingParamList1["am22"][[1]][[condition]] #iq.m - iq.f

	#VT
	f11 <- startingParamList1["f11"][[1]][[condition]] # regression of offspring trait 1 F on parental trait 1
	f12 <- startingParamList1["f12"][[1]][[condition]] # regression of offspring trait 1 F on parental trait 2
	f21 <- startingParamList1["f21"][[1]][[condition]] # regression of offspring trait 2 F on parental trait 1
	f22 <- startingParamList1["f22"][[1]][[condition]] # regression of offspring trait 2 F on parental trait 2

	#E
	re <- startingParamList1["re"][[1]][[condition]] #environmental CORRELATION between trait 1 & trait 2

	# IMPLIED variables
	#This section just converts the above inputs into matrices in our algebra

	#Create lower delta matrix (d stands for delta)
	(vg.obs1 <- vg1*(1-prop.h2.latent1))
	(vg.obs2 <- vg2*(1-prop.h2.latent2))
	(covg.obs <- rg*sqrt(vg.obs1)*sqrt(vg.obs2))
	d11 <- sqrt(vg.obs1)
	#d21 <- covg.obs/d11
	d21 <- 0 #new way
	d22 <- sqrt(vg.obs2 - d21^2)
	(delta.mat <- matrix(c(d11,0,d21,d22),byrow=T,nrow=2))
	(vg.obs <- delta.mat %*% k2.matrix %*% t(delta.mat)) #Check
	matrix(c(vg.obs1,covg.obs,covg.obs,vg.obs2),nrow=2) #Check

	#Create lower a matrix
	(vg.lat1 <- vg1*(prop.h2.latent1))
	(vg.lat2 <- vg2*(prop.h2.latent2))
	(covg.lat <- rg*sqrt(vg.lat1)*sqrt(vg.lat2))
	a11 <- sqrt(vg.lat1)
	#a21 <- covg.lat/a11
	a21 <- 0 #new way
	a22 <- sqrt(vg.lat2 - a21^2)
	(a.mat <- matrix(c(a11,0,a21,a22),byrow=T,nrow=2))
	(vg.lat <- a.mat %*% k2.matrix %*% t(a.mat)) #Check
	matrix(c(vg.lat1,covg.lat,covg.lat,vg.lat2),nrow=2) #Check

	#Find the total genetic covariance
	#(covg <- vg.obs[1,2] + vg.lat[1,2])
	(covg.mat <- vg.obs + vg.lat)
	a.mat %*% k2.matrix %*% t(a.mat) + delta.mat %*% k2.matrix %*% t(delta.mat) #check

	#Create E lower matrix
	ve1 <- 1 - vg1
	ve2 <- 1 - vg2
	(cove <- re*sqrt(ve1*ve2))
	(cove.mat <- matrix(c(ve1,cove,cove,ve2),nrow=2,byrow=T))

	#Create var-covar(Y) @t0. NOTE: we define the base population as the population before any AM or VT has occurred.
	(COVY <- covg.mat+cove.mat)
	a.mat %*% k2.matrix %*% t(a.mat) + delta.mat %*% k2.matrix %*% t(delta.mat) + cove.mat  #Check; these two should have 1 on the diagonals and the off-diags should be exactly the same if we've done everything above correctly.
	#Note that our actual model estimates k and j separately, and thus our model does not make the assumption made in this simulation that the observed and latent rg at t0 are the same (or does it? do we actually estimate j12, or do we set it equal to k12?)


	#Note: we will add influences of AM and VT AFTER time 0
	#AM - make sure to change am.list if you want AM to change across generations
	(mate.cor.mat <- matrix(c(am11,am12,am21,am22),nrow=2,byrow=T))
	#am.list <- rep(list(mate.cor.mat),num.gen+1)

	#VF
	(f.mat <- matrix(c(f11,f12,f21,f22),nrow=2,byrow=T))
	(covf.mat <- 2 * (f.mat %*% COVY %*% t(f.mat)))
    

 
    a.t0 <- a.mat
    delta.t0 <- delta.mat
    j.t0 <- k2.matrix*.5
    k.t0 <- k2.matrix*.5
    f.t0 <- f.mat
    rmate.t0 <- mate.cor.mat
    covE.t0 <- cove.mat
    #rmate.t0 <- obs.cor

    #Define and initialize at t0 parameters that evolve
    exp.gc <- exp.hc <- exp.ic <- exp.gt <- exp.ht <- exp.itlo <- exp.itol <- exp.w <- exp.v <- exp.VY <- exp.VY.diagonal <- exp.corVY<- exp.mu <- exp.VF <- exp.Omega <- exp.Gamma <- exp.thetaT <- exp.thetaNT <- exp.thetaLT <- exp.thetaLNT <- mate.cov <- mate.cor <- exp.VGO <- exp.VGL <- exp.COVLO <- vector(mode="list",length=gens-1)
    #initialize parameters that begin as 0's in base population (before VT and AM)
    exp.VGO[[1]] <- exp.VGL[[1]] <- exp.COVLO[[1]] <- exp.gc[[1]] <- exp.hc[[1]] <- exp.ic[[1]] <- exp.w[[1]] <- exp.v[[1]] <- matrix(0,nrow=2,ncol=2)
    # define the true 4X4 covariance matrix for Vy at t0
    exp.covY.full <- exp.corY.full <- exp.covY.full.diagonal <- vector(mode="list",length=gens-1)
    #initialize parameters that are non-zero in base population (think of these as the first parents to pass on VT and engage in AM)
    (exp.VY[[1]] <- COVY)
    (exp.VY.diagonal[[1]] <- diag(diag(exp.VY[[1]])))
    (exp.VF[[1]] <- covf.mat)
    (exp.mu[[1]] <- solve(COVY) %*% rmate.t0 %*% solve(t(COVY)))
    (mate.cov[[1]] <- COVY %*% exp.mu[[1]] %*% COVY )
    (exp.covY.full[[1]] <- rbind(cbind(exp.VY[[1]], mate.cov[[1]]),cbind(mate.cov[[1]], exp.VY[[1]])))
    (exp.corY.full[[1]] <- cov2cor(exp.covY.full[[1]]))
    mate.cor[[1]] <- mate.cov[[1]]/COVY
    mate.cor[[1]][1,2] <- mate.cov[[1]][1,2]/sqrt(COVY[1,1]*COVY[2,2])
    mate.cor[[1]][2,1] <- mate.cov[[1]][2,1]/sqrt(COVY[1,1]*COVY[2,2])
    
```

# third chunk: run the iterative math and print out parameters each generation
```{r iterativeMath}
# run the loop of iterative math for the specified number of generations
# add more cat() if you want to print out more matrices
for (it in 2:gens){
        (exp.Omega[[it]] <- 2*delta.t0%*%exp.gc[[it-1]] + delta.t0 %*% k.t0 + .5*exp.w[[it-1]] + 2*a.t0%*%exp.ic[[it-1]])
        cat("Omega",it,"\n",exp.Omega[[it]],"\n")

        (exp.Gamma[[it]] <- 2*a.t0%*%exp.hc[[it-1]] + 2*delta.t0%*%t(exp.ic[[it-1]]) + a.t0%*%j.t0 + .5*exp.v[[it-1]])
        cat("Gamma",it,"\n",exp.Gamma[[it]],"\n")

        (exp.VY[[it]] <- 2 * delta.t0 %*% t(exp.Omega[[it]]) + 2 * a.t0 %*% t(exp.Gamma[[it]]) + exp.w[[it-1]] %*% t(delta.t0) + exp.v[[it-1]] %*% t(a.t0) + exp.VF[[it-1]] + covE.t0 )
        cat("VY",it,"\n")
        print(exp.VY[[it]])

        exp.VY.diagonal[[it]] <- exp.VY[[it]]
        exp.VY.diagonal[[it]][1,2] <- 0
        exp.VY.diagonal[[it]][2,1] <- 0
        exp.covY.full.diagonal[[it]] <- cbind(rbind(exp.VY.diagonal[[it]],matrix(0,2,2)),rbind(matrix(0,2,2),exp.VY.diagonal[[it]]))
        exp.corVY[[it]] <- cov2cor(exp.VY[[it]])
        exp.corY.full[[it]] <- exp.corY.full[[1]]
        exp.corY.full[[it]][1,2] <- exp.corVY[[it]][1,2]
        exp.corY.full[[it]][2,1] <- exp.corVY[[it]][2,1]
        exp.corY.full[[it]][3,4] <- exp.corVY[[it]][1,2]
        exp.corY.full[[it]][4,3] <- exp.corVY[[it]][2,1]
        #exp.covY.full[[it]] <- cor2cov(exp.corY.full[[it]],exp.VY[[it]][1,1],exp.VY[[it]][2,2],exp.VY[[it]][1,1],exp.VY[[it]][2,2]) 
        exp.covY.full[[it]] <- sqrt(exp.covY.full.diagonal[[it]]) %*% exp.corY.full[[it]] %*% sqrt(exp.covY.full.diagonal[[it]])
        mate.cov[[it]] <- exp.covY.full[[it]][3:4,1:2] ## this is the mate.cov fixing mate.cor
        mate.cor[[it]] <- exp.corY.full[[1]][3:4,1:2]
        (exp.mu[[it]] <- solve(exp.VY[[it]]) %*% mate.cov[[it]] %*% solve(t(exp.VY[[it]]))) 

        (exp.gt[[it]] <- t(exp.Omega[[it]])%*%exp.mu[[it]]%*%exp.Omega[[it]])
        cat("gt",it,"\n",exp.gt[[it]],"\n")

        (exp.gc[[it]] <- .5*(exp.gt[[it]] + t(exp.gt[[it]])))
        cat("gc",it,"\n",exp.gc[[it]],"\n")     

        (exp.ht[[it]] <- t(exp.Gamma[[it]])%*%exp.mu[[it]]%*%exp.Gamma[[it]])
        cat("ht",it,"\n",exp.ht[[it]],"\n")

        (exp.hc[[it]] <- .5*(exp.ht[[it]] + t(exp.ht[[it]])))  
        cat("hc",it,"\n",exp.hc[[it]],"\n")

        (exp.w[[it]] <- 2*f.t0%*%exp.Omega[[it]] + f.t0%*%exp.VY[[it]]%*%exp.mu[[it]]%*%exp.Omega[[it]] + f.t0%*%exp.VY[[it]]%*%t(exp.mu[[it]])%*%exp.Omega[[it]])
        cat("w",it, "\n")
        print(exp.w[[it]])

        (exp.v[[it]] <- 2*f.t0%*%exp.Gamma[[it]] + f.t0%*%exp.VY[[it]]%*%exp.mu[[it]]%*%exp.Gamma[[it]] + f.t0%*%exp.VY[[it]]%*%t(exp.mu[[it]])%*%exp.Gamma[[it]])  
        cat("v",it, "\n")
        print(exp.v[[it]])

        (exp.VF[[it]] <- 2*f.t0%*%exp.VY[[it]]%*%t(f.t0) + f.t0%*%exp.VY[[it]]%*%exp.mu[[it]]%*%exp.VY[[it]]%*%t(f.t0) +  f.t0%*%exp.VY[[it]]%*%t(exp.mu[[it]])%*%exp.VY[[it]]%*%t(f.t0))
        cat("VF",it,"\n")
        print(exp.VF[[it]])

        (exp.itlo[[it]] <- t(exp.Gamma[[it]])%*%exp.mu[[it]]%*%exp.Omega[[it]])
        
        (exp.itol[[it]] <- t(exp.Omega[[it]])%*%exp.mu[[it]]%*%exp.Gamma[[it]])

        (exp.ic[[it]] <- .5*(exp.itol[[it]] + t(exp.itlo[[it]])))

        (exp.VGO[[it]] <- 2*delta.t0%*%k.t0%*%t(delta.t0) + 4*delta.t0%*%exp.gc[[it]]%*%t(delta.t0))

        (exp.VGL[[it]] <- 2*a.t0%*%j.t0%*%t(a.t0) + 4*a.t0%*%exp.hc[[it]]%*%t(a.t0))
        
        (exp.COVLO[[it]] <- 4*delta.t0%*%t(exp.ic[[it]])%*%t(a.t0) + 4*a.t0%*%exp.ic[[it]]%*%t(delta.t0))
}

```

# fourth chunk: play with the constraints of g,h,w,v
```{r checkConstraints}
# the package expm to check the logarithm of the constraints
library(expm)
# this chunk is to check the constraints of h,g,w,v
# get the values of delta, a, j and k
delta <- delta.t0
a <- a.t0
k <- k.t0
j <- j.t0
# get the values of h, g, w, v of the final generation
hc <- exp.hc[[gens]]
gc <- exp.gc[[gens]]
w <- exp.w[[gens]]
v <- exp.v[[gens]]
# three ways of constraining h, g
# devision
gc/(2* delta %*% k %*% t(delta))
hc/(2* a %*% j %*% t(a))
# inverse
gc * solve(2* delta %*% k %*% t(delta))
hc * solve(2* a %*% j %*% t(a))
# cholesky
t(chol(solve(2* delta %*% k %*% t(delta)))) %*% gc %*% chol(solve(2* delta %*% k %*% t(delta)))
t(chol(solve(2* a %*% j %*% t(a)))) %*% hc %*% chol(solve(2* a %*% j %*% t(a)))
# log
logm(gc) - logm(2* delta %*% k %*% t(delta))
logm(hc) - logm(2* a %*% j %*% t(a))

# three ways of constraining w, v
# devision
w/(2* delta %*% k %*% t(delta))
v/(2* a %*% j %*% t(a))
# inverse
v %*% solve(2* a %*% j %*% t(a))
w %*% solve(2* delta %*% k %*% t(delta))
# cholesky
t(chol(solve(sqrt(2* delta %*% k %*% t(delta))))) %*% (w) %*% chol(solve(sqrt(2* delta %*% k %*% t(delta))))
t(chol(solve(sqrt(2* a %*% j %*% t(a))))) %*% (v) %*% chol(solve(sqrt(2* a %*% j %*% t(a))))
logm(w) - logm(2* delta %*% k %*% t(delta))
logm(v) - logm(2* a %*% j %*% t(a))


```